---
title: "R Reference Sheets"
output: 
  flexdashboard::flex_dashboard:
    theme: cerulean
    highlight: tango
    orientation: rows # Determines whether level 2 headings are treated as dashboard rows or dashboard columns.
    vertical_layout: fill
    css: custom.css
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, fig.width = 4, fig.height =3)
```

Basic usage of R
===================================== 

Row {data-height=250}
-----------------------------------------------------------------------

### Mathematical Operators
```{r, eval=FALSE}
# Addition
5 + 5   # [1] 10
# Subtraction
10 - 4  # [1] 6
# Multiplication
6 * 7   # [1] 42
# Division
8 / 2   # [1] 4

# Exponent
2 ^ 5   # [1] 32
2 ** 5  # [1] 32
```

### Logical Operators
```{r, eval=FALSE}
# Equal to
5 == 5     # [1] TRUE
# Not equal to
10 != 5    # [1] TRUE 
# Greater (less) than 
10 > 2     # [1] TRUE
10 < 4     # [1] FALSE
# Greater (less) than or equal to
10 >= 10   # [1] TRUE
10 <= 11   # [1] TRUE
# Not greater (less) than 
!(5 > 3)  # [1] FALSE
!(10 < 4)  # [1] TRUE
```


### Defining variables
```{r, eval=FALSE}
# Define numeric variable `x` with a value of 7
x <- 7
# Define numeric variable `y` with a value of 7.25
7.25 -> y

# Examples of defining some character variables
best_food  <- "ice cream"
worst_food <- 'zucchini'

# Output saved to variable.
math_result <- 5*5 #
```
*Remember: You must __print__ the variable to see its value. Defining a variable does not print the variable you its value automatically!*

Row {data-height=500}
-----------------------------------------------------------------------

### Arrays

```{r, eval=FALSE}
# Define numeric array variable `x`, using the c() function
x <- c(0, 1, 1, 2, 3, 5, 8, 13)

# Perform operations, run functions on arrays:
x * 2    # [1] 0  2  2  4  6  10  16  26
x >= 5   # [1] FALSE FALSE FALSE FALSE FALSE TRUE TRUE TRUE
log(x)   # [1] -Inf 0.00 0.00 0.69 1.09 1.60 2.07 2.56

# Define character array variable `y` 
y <- c("a", "b", "c", "d", "e", "f")

# Define array variable `z` as a sequence 1, 2, 3, ..., 98, 99, 100
z <- 1:100
```


```{r, eval=F}
# Compute summary statistics from arrays.
bird_weights <- c(5.1, 4.7, 3.6, 4.2, 3.7, 6, 5.8, 6.1, 5.5, 3.8, 6.6, 4.3)

mean(bird_weights)   # [1] 4.95
median(bird_weights) # [1] 4.9
min(bird_weights)    # [1] 3.6
max(bird_weights)    # [1] 6.6
sd(bird_weights)     # [1] 1.042288
summary(bird_weights) # Output below:
#   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   3.60    4.10    4.90    4.95    5.85    6.60 

# If there are NAs in the data, use argument `na.rm=TRUE` to ignore in calculations:
bird_weights_missing <- c(5.1, 4.7, 3.6, NA)
mean(bird_weights_missing)                # [1] NA    # womp :(
mean(bird_weights_missing, na.rm=TRUE)    # [1] 4.466667  
```

### Data types in R

<!--
+ Use the function `class()` to determine the type of a variable
+ Use the function `as.TYPE()` to coerce a variable to change type
  + `TYPE` is one of `numeric`, `character()`, `logical()` or `factor()`
+ Use the function `factor(..., levels=c(), labels=c())` to change factor order or labels
-->
```{r, eval=FALSE}
## Examples with numeric data variable
x <- 10              # (no output - just defines x!)
class(x)             # "numeric"
as.character(x)      # "10" . Does NOT change value of x
as.logical(x)        # TRUE . Does NOT change value of x
x <- as.character(x) # Actually changes value of `x` to "10"
```
```{r, eval=FALSE}
## Examples with character data variable
a_fruit <- "apple"
class(a_fruit)       # "character"
as.numeric(a_fruit)  # NA (no way to convert "apple" to a number!)
```
```{r, eval=FALSE}
## Examples with factor data variable
several_fruits <- c("apple", "apple", "banana", "cherry", "papaya", "papaya", "papaya") 
several_fruits <- as.factor(several_fruits) # Coerce into a factor
class(several_fruits)   # "factor"
levels(several_fruits)  # [1] "apple"  "banana" "cherry" "papaya"

# Use function `factor()` to RE-ORDER the levels
several_fruits2 <- factor(several_fruits,  
                          levels = c("papaya", "apple", "cherry", "banana"))
levels(several_fruits2)  # [1] "papaya" "apple"  "cherry" "banana"

# re-order levels AND change the labels
several_fruits3 <- factor(several_fruits,  
                          levels = c("papaya", "apple", "cherry", "banana"),
                          labels = c("Papaya", "Apple", "Cherry", "Banana"))
levels(several_fruits3)  # [1] "Papaya" "Apple"  "Cherry" "Banana"

```


Data frames and functions  {data-orientation=columns}
=====================================


Column
----------------------------------------------------------

### Miscellaneous functions


```{r, eval=FALSE}
# Useful functions for numbers
x <- 100

# Natural log by default. Second argument specifies base
log(x)       # [1] 4.60517 
log(x, 10)   # [1] 2

# Square root
sqrt(x)      # [1] 10
```

```{r, eval=FALSE}
# Various ways to round
x <- 100.679274
round(x)       # [1] 101
round(x, 2)    # [1] 100.68
floor(x)       # [1] 100 . Round down to nearest whole number
ceiling(x)     # [1] 101 . Round up to nearest whole numer 
```

```{r, eval=FALSE}
# Use functions for arrays
x <- c(0, 1, 1, 2, 3, 5, 8, 13)

# Number of elements in an array
length(x)       # [1] 8
# Sum of elements in array (if numeric or logical!)
sum(x)          # [1] 33

logical_array <- c(TRUE, FALSE, FALSE, TRUE, TRUE) # TRUE=1; FALSE=0
sum(logical_array)  # [1] 3

# all() asks: Are ALL values TRUE for the condition?
all(x > 5)      # [1] FALSE
# any() asks: Are ANY values TRUE for the condition?
any(x > 5)      # [1] TRUE

# Find all unique values 
unique(x)       # [1] 0  1  2  3  5  8 13
```

```{r, eval=FALSE}
# Number of characters in a character variable
x <- "pineapple"
nchar(x)      # [1] 9
```

```{r, eval=FALSE}
# Check for NA's in data
x <- "pineapple"
is.na(x)     # [1] FALSE

x <- c(1, 4, NA, 10, 24, NA, 13)
is.na(x)     # [1] FALSE FALSE TRUE FALSE FALSE TRUE FALSE
```




Column
----------------------------------------------------------


### Exploring data frames


```{r, eval=FALSE}
# View first 6 rows
head(iris)
#  Sepal.Length Sepal.Width Petal.Length Petal.Width Species
#1          5.1         3.5          1.4         0.2  setosa
#2          4.9         3.0          1.4         0.2  setosa
#3          4.7         3.2          1.3         0.2  setosa
#4          4.6         3.1          1.5         0.2  setosa
#5          5.0         3.6          1.4         0.2  setosa
#6          5.4         3.9          1.7         0.4  setosa

# View first 2 rows onlu
head(iris, 2)
#  Sepal.Length Sepal.Width Petal.Length Petal.Width Species
#1          5.1         3.5          1.4         0.2  setosa
#2          4.9         3.0          1.4         0.2  setosa
```

*Within the `RStudio` application, you can use `View(iris)`. This `View()` function does not work anywhere else!*

```{r, eval=FALSE}
# Extract a column as an array with the dollar sign $
iris$Sepal.Length
# [1] 5.1 4.9 4.7 4.6 5.0 5.4 4.6 5.0 4.4 4.9 5.4 4.8 4.8 4.3 5.8 5.7 5.4 5.1
# [19] 5.7 5.1 5.4 5.1 4.6 5.1 4.8 5.0 5.0 5.2 5.2 4.7 4.8 5.4 5.2 5.5 4.9 5.0
# [37] 5.5 4.9 4.4 5.1 5.0 4.5 4.4 5.0 5.1 4.8 5.1 4.6 5.3 5.0 7.0 6.4 6.9 5.5
# [55] 6.5 5.7 6.3 4.9 6.6 5.2 5.0 5.9 6.0 6.1 5.6 6.7 5.6 5.8 6.2 5.6 5.9 6.1
# [73] 6.3 6.1 6.4 6.6 6.8 6.7 6.0 5.7 5.5 5.5 5.8 6.0 5.4 6.0 6.7 6.3 5.6 5.5
# [91] 5.5 6.1 5.8 5.0 5.6 5.7 5.7 6.2 5.1 5.7 6.3 5.8 7.1 6.3 6.5 7.6 4.9 7.3
#[109] 6.7 7.2 6.5 6.4 6.8 5.7 5.8 6.4 6.5 7.7 7.7 6.0 6.9 5.6 7.7 6.3 6.7 7.2
#[127] 6.2 6.1 6.4 7.2 7.4 7.9 6.4 6.3 6.1 7.7 6.3 6.4 6.0 6.9 6.7 6.9 5.8 6.8
#[145] 6.7 6.7 6.3 6.5 6.2 5.9
```


```{r, eval=FALSE}
# Ask how many rows, columns in the data frame
nrow(iris)   # [1] 150
ncol(iris)   # [1] 5
```

```{r, eval=FALSE}
# Summarize the whole data frame
summary(iris)
#  Sepal.Length    Sepal.Width     Petal.Length    Petal.Width   
# Min.   :4.300   Min.   :2.000   Min.   :1.000   Min.   :0.100  
# 1st Qu.:5.100   1st Qu.:2.800   1st Qu.:1.600   1st Qu.:0.300  
# Median :5.800   Median :3.000   Median :4.350   Median :1.300  
# Mean   :5.843   Mean   :3.057   Mean   :3.758   Mean   :1.199  
# 3rd Qu.:6.400   3rd Qu.:3.300   3rd Qu.:5.100   3rd Qu.:1.800  
# Max.   :7.900   Max.   :4.400   Max.   :6.900   Max.   :2.500  
#       Species  
# setosa    :50  
# versicolor:50  
# virginica :50  
```





`dplyr` {data-orientation=columns}
===================================== 

Column 
-----------------------------------------------------------------------

### Using the pipe operator `%>%`
```{r, eval=FALSE}
# first_input_argument %>%
#   function_name(second_input_argument)

# Equivalent to log(5)
5 %>% log() # [1] 1.609438

# Equivalent to log(5, 10)
5 %>% log(10) # [1] 0.69897
```


### First, look at your data
```{r, eval=FALSE}
iris
# A tibble: 150 x 5
#   Sepal.Length Sepal.Width Petal.Length Petal.Width Species
#          <dbl>       <dbl>        <dbl>       <dbl> <fct>  
# 1          5.1         3.5          1.4         0.2 setosa 
# 2          4.9         3            1.4         0.2 setosa 
# 3          4.7         3.2          1.3         0.2 setosa 
# 4          4.6         3.1          1.5         0.2 setosa 
# 5          5           3.6          1.4         0.2 setosa 
# 6          5.4         3.9          1.7         0.4 setosa 
# 7          4.6         3.4          1.4         0.3 setosa 
# 8          5           3.4          1.5         0.2 setosa 
# 9          4.4         2.9          1.4         0.2 setosa 
#10          4.9         3.1          1.5         0.1 setosa 
```

```{r, eval=FALSE}
dplyr::glimpse(iris)
#Rows: 150
#Columns: 5
#$ Sepal.Length <dbl> 5.1, 4.9, 4.7, 4.6, 5.0, 5.4, 4.6, 5.0, 4.4, 4.9, 5.4, 4.8, …
#$ Sepal.Width  <dbl> 3.5, 3.0, 3.2, 3.1, 3.6, 3.9, 3.4, 3.4, 2.9, 3.1, 3.7, 3.4, …
#$ Petal.Length <dbl> 1.4, 1.4, 1.3, 1.5, 1.4, 1.7, 1.4, 1.5, 1.4, 1.5, 1.5, 1.6, …
#$ Petal.Width  <dbl> 0.2, 0.2, 0.2, 0.2, 0.2, 0.4, 0.3, 0.2, 0.2, 0.1, 0.2, 0.2, …
#$ Species      <fct> setosa, setosa, setosa, setosa, setosa, setosa, setosa, seto…
```

<!--
### Examples: A full wrangle

*How many irises of each species have a sepal width larger than 4 cm?* **Only 3 setosa irises!**

```{r}
iris %>%
  # Create species grouping
  dplyr::group_by(Species) %>%
  # Subset to only petal lengths of interest
  dplyr::filter(Sepal.Width > 4) %>%
  # Count the number of rows per group
  dplyr::tally()
```

-->

Column
-------------------------------------------

### Subset columns with `dplyr::select()`
```{r, eval=FALSE}
# Only keep Sepal.Length, Sepal.Width
iris %>%
  dplyr::select(Sepal.Length, Sepal.Width)

# Remove columns Sepal.Length, Sepal.Width
iris %>% 
  dplyr::select(-Sepal.Length, -Sepal.Width)
```

### Rename columns with `dplyr::rename()`
```{r, eval=FALSE}
# Change "Sepal.Length" to sepal_length
iris %>%
  dplyr::rename(sepal_length = Sepal.Length)
```


### Create new columns with `dplyr::mutate()`
```{r, eval=FALSE}
# Create a column called "flower" whose value is simply "iris"
iris %>%
  dplyr::mutate(flower = "iris")

# Create a column sepal_area as the product Sepal.Length*Sepal.Width
iris %>%
  dplyr::mutate(sepal_area = Sepal.Length*Sepal.Width)
```



Column
----------------------------------------


### Subset rows with `dplyr::filter()`
```{r, eval=FALSE}
# Remove all versicolor Species
iris %>%
  dplyr::filter(Species != "versicolor")

# Keep only rows with Sepal.Lengths larger than 5 cm AND Species is setosa
iris %>% 
  dplyr::filter(Sepal.Length > 5, Species == "setosa")
```


### Arrange rows with `dplyr::arrange()`
```{r, eval=FALSE}
# Arrange in ascending order of Sepal.Length values
iris %>%
  dplyr::arrange(Sepal.Length)

# Arrange in descending order of Sepal.Length values
iris %>%
  dplyr::arrange(desc(Sepal.Length))
```





`ggplot2`
===================================== 


`tidyr`
===================================== 


Miscellaneous `tidyverse`
===================================== 

